Supplementary material for: non-coding RNAs of bird genomes

### 2014-05-23 paper/manuscript.tex
Addressing Reviewer comments.

The conservation of these ``human'' lncRNAs among birds suggests they may 
also be functional in birds but what these
functions is not immediately obvious. For example, PART1 and
PCA3 are both described as prostate-specific lncRNAs that play a role
in the human androgen-receptor pathway
\cite{Bussemakers:1999,Lin:2000,Ferreira:2012}. Birds lack a prostate
but both males and females express the androgen receptor (AR or NR3C4) in gonadal
and non- gonadal tissue
\cite{Yoshimura:1993,Veney:2004,Fuxjager:2012,Leska:2012}. Thus, we
postulate that PART1 and PCA3 also play a role in the
androgen-receptor pathway in birds but whether the expression of these
lncRNAs are tissue specific is unknown at present.

@@

Of the other classes of RNAs, none showed an unexpected degree of
conservation or expansion within the avian lineage. The only exception
being the snoRNA, SNORD93 which has 92 copies in the tinamou
genome, whereas it only has 1-2 copies in all the other vertebrate
genomes.

@@

The overall reduction in avian genomic size has been extensively
discussed elsewhere \cite{Organ:2007}. Unsurprisingly, this reduction
is reflected in the copy-number of ncRNA genes. Some of the most
dramatic examples are the transfer RNAs and pseudogenes which average
$\sim900$ and $\sim580$ copies in the human, turtle and alligator
genomes, the average copy numbers of these drop to $\sim280$ and
$\sim100$ copies in the avian genomes. In addition to reduction in
copy-number, the complete absence of several otherwise ubiquitous
vertebrate ncRNAs in the avian lineage are indicative of genuine gene
loss.

@@

Vertebrate Y RNAs typically form a cluster comprising four
well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
\cite{Mosig:07a} we find that the Y5 paralog family is absent from all
bird genomes, while it is still present in both alligator and turtle,
see Supplemental Figure~4. Within the avian lineage, we find a
conserved Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most
cases consistent with breaks in the available sequence assemblies.  In
several genomes we observe one or a few additional Y RNA homologs
unlinked to the canonical Y RNA cluster. These sequences can be
identified unambiguously as derived members of one of the three
ancestral paralog groups, they almost always fit less well to the
consensus (as measured by the CM bit score of paralog group specific
covariance models) than the paralog linked to cluster, and there is no
indication that any of these additional copies is evolutionarily
conserved over longer time scales. We therefore suggest that most or
all of these interspersed copies are in fact pseudogenes.

@@

Heatmaps showing the presence/absence and approximate genomic copy
number of ``unusually, well conserved RNAs'' (particularly the
lncRNAs) on the left and families that have been identified as
RNA losses, divergence or missing data. In several cases
functionally related families have also been included, e.g. the RNA
components of the major and minor spliceosomes: U1, U2, U4, U5 and U6;
and U11, U12, U4atac, U5 and U6atac, respectively.\label{fig:1}

>>>>

The conservation of these ``human'' lncRNAs among birds suggests they
may also be functional in birds. But what these functions are, is not
immediately obvious. For example, PART1 and PCA3 are both described as
prostate-specific lncRNAs that play a role in the human
androgen-receptor pathway
\cite{Bussemakers:1999,Lin:2000,Ferreira:2012}. Birds lack a prostate
but both males and females express the androgen receptor (AR or NR3C4)
in gonadal and non- gonadal tissue
\cite{Yoshimura:1993,Veney:2004,Fuxjager:2012,Leska:2012}. Thus, we
postulate that PART1 and PCA3 also play a role in the
androgen-receptor pathway in birds but whether the expression of these
lncRNAs are tissue specific is unknown at present.

@@

Of the other classes of RNAs, none showed an unexpected degree of
conservation or expansion within the avian lineage. The only exception
being the snoRNA, SNORD93. SNORD93 has 92 copies in the tinamou
genome, whereas it only has 1-2 copies in all the other vertebrate
genomes.

@@

The overall reduction in avian genomic size has been extensively
discussed elsewhere \cite{Organ:2007}. Unsurprisingly, this reduction
is reflected in the copy-number of ncRNA genes. Some of the most
dramatic examples are the transfer RNAs and pseudogenes which average
$\sim900$ and $\sim580$ copies in the human, turtle and alligator
genomes, the average copy-numbers of these drop to $\sim280$ and
$\sim100$ copies in the avian genomes. In addition to reduction in
copy-number, the absence of several, otherwise ubiquitous vertebrate
ncRNAs, in the avian lineage are suggestive of genuine gene loss.

@@

Vertebrate Y RNAs typically form a cluster comprising four
well-defined paralog groups Y1, Y3, Y4, and Y5. In line with
\cite{Mosig:07a} we find that the Y5 paralog family is absent from all
bird genomes, while it is still present in both alligator and turtle,
see Supplemental Figure~4. Within the avian lineage, we find a
conserved Y4-Y3-Y1 cluster. Apparently, broken-up clusters are in most
cases consistent with breaks (e.g. ends of contigs) in the available
sequence assemblies.  In several genomes we observe one or a few
additional Y RNA homologs unlinked to the canonical Y RNA
cluster. These sequences can be identified unambiguously as derived
members of one of the three ancestral paralog groups, they almost
always fit less well to the consensus (as measured by the CM bit score
of paralog group specific covariance models) than the paralog linked
to cluster, and there is no indication that any of these additional
copies is evolutionarily conserved over longer time scales. We
therefore suggest that most or all of these interspersed copies are in
fact pseudogenes.

@@

Heatmaps showing the presence/absence and approximate genomic
copy-number of ``unusually, well conserved RNAs'' (particularly the
lncRNAs) on the left and families that have been identified as RNA
losses, divergence or missing data. In several cases functionally
related families have also been included, e.g. the RNA components of
the major and minor spliceosomes: U1, U2, U4, U5 and U6; and U11, U12,
U4atac, U5 and U6atac, respectively.\label{fig:1}

### 2014-05-23 paper/supplementary-material.tex
Addressing Reviewer comments.

A number of the classic RNAs are incorporated into RNA-protein
complexes (RNPs) involved in core cellular processes. An example of
this are the spliceosomal RNAs. Based upon the presence/absence
patterns of the major spliceosomal RNAs they are all well represented
in these genome sequences. Except for the U4 RNA in cormorant and the
U5 RNA in the bee eater. Again these two genomes are low coverage
suggesting these genes weren't captured in the current assembly. The
minor spliceosomal RNAs are more interesting, the U4atac and U11
snRNAs show widespread patterns of loss, even in some of the high
coverage genomes. These RNAs are frequently missed in bioinformatic
screens. Indicating either frequent loss \cite{Davila_Lopez:2008} or
sequences that have diverged beyond the ability of detection by
covariance models \cite{Marz:2008}.

@@

The first cluster is found at chr11:62620797-62622484 on the human
genome (hg19) and contains SNORD27, SNORD29 and SNORD31 of the
human-yeast conserved snoRNAs. These snoRNAs are located in the
inside-out gene SNHG1 which hosts a total of eight C/D box snoRNAs:
SNORD25, SNORD26, SNORD27, SNORD28, SNORD29, SNORD22, SNORD30 and
SNORD31 \cite{Tycowski:1996}. Each of which are also found in the
alligator and turtle genomes within a 3-4 KB locus, yet these have
largely been lost in the birds. Notable exceptions are five of the
eight are located in the tinamou genome, these are located on the same
scaffold and are within 2 KB of each other. Implying that SNHG1 is
conserved in the tinamou. Loci with four of the eight snoRNAs can be
found in zebrafinch, ground-finch, and bald eagle. Still, three of the
eight are located in the ostrich, crow, and cockoo genomes, again
within 2 KB of each other on the same scaffolds. This complex pattern
of loss could be attributed to many different models, e.g. multiple
losses in birds, poor homology modelling or incomplete genome
sequences.

>>>>

A number of the classic RNAs are incorporated into RNA-protein
complexes (RNPs) involved in core cellular processes. An example of
this are the spliceosomal RNAs. Based upon the presence/absence
patterns of the major spliceosomal RNAs they are all well represented
in these genome sequences. The exceptions to this observation are the
U4 RNA in cormorant and the U5 RNA in the bee eater which are both
missing. These two genomes are low coverage, suggesting these
genes weren't captured in the current assembly. The minor spliceosomal
RNAs are more interesting, the U4atac and U11 snRNAs show widespread
patterns of loss, even in some of the high coverage genomes. These
RNAs are frequently missed in bioinformatic screens. Indicating either
frequent loss \cite{Davila_Lopez:2008} or sequences that have diverged
beyond the ability of detection by covariance models \cite{Marz:2008}.

@@

The first cluster is found at chr11:62620797-62622484 on the human
genome (hg19) and contains SNORD27, SNORD29 and SNORD31 of the
human-yeast conserved snoRNAs. These snoRNAs are located in the
inside-out gene SNHG1 which hosts a total of eight C/D box snoRNAs:
SNORD25, SNORD26, SNORD27, SNORD28, SNORD29, SNORD22, SNORD30 and
SNORD31 \cite{Tycowski:1996}. Each of which are also found in the
alligator and turtle genomes within a 3-4 KB locus, yet these have
largely been lost in the birds. Notable exceptions are five of the
eight are located in the tinamou genome, these are located on the same
scaffold and are within 2 KB of each other. This implies that SNHG1 is
conserved in the tinamou. Loci with four of the eight snoRNAs can be
found in zebrafinch, ground-finch, and bald eagle. Still, three of the
eight are located in the ostrich, crow, and cockoo genomes, again
within 2 KB of each other on the same scaffolds. This complex pattern
of loss could be attributed to many different models, e.g. multiple
losses in birds, poor homology modelling or incomplete genome
sequences.
